HyperLogLog: Exponentially Bad in Adversarial Settings

### 2021-04-30 docs/HLL_exp_bad/main.tex
fixing up proof

\todo[inline]{This ends up being pretty harsh on this NSDI paper, but hey.}

@@

\subsection{Intuition for Formal Security Analysis for S2, S3, S4}

@@

In the ideal world ($b=1$), ${\mathcal{A}}$ is replaced by a simulator ${\mathcal{S}}$ with blackbox oracle access to ${\mathcal{A}}$ (that is, ${\mathcal{S}}$ can run ${\mathcal{A}}$ in a blackbox manner, receiving its oracle queries and providing responses to ${\mathcal{A}}$ as it sees fit). In particular, ${\mathcal{S}}$ receives parameters $n$, $\ell$, $\alpha_n$ from ${\mathcal{A}}$'s \texttt{init($\cdot$)} call. Finally, ${\mathcal{S}}$ receives $\textsf{out}_{\mathcal{A}}$ and completes its execution by outputting its own output $\textsf{out}_{\mathcal{S}}$.

Importantly, we will specify a construction of ${\mathcal{S}}$ that handles any of ${\mathcal{A}}$'s \texttt{insert($x$)} queries using \emph{random sampling}, that is, by setting $y \leftarrow_{\$} \{0,1\}^{n+\ell}$ instead of $y: = F_K(x)$ and then executing the rest of the insertion process as in Figure~\ref{fig:hll}. In essence, our ${\mathcal{S}}$ will build its own ``shadow'' HLL in which all insertion queries are handled by random sampling. ${\mathcal{S}}$ then uses the shadow HLL for handling ${\mathcal{A}}$'s \texttt{read} and \texttt{corrupt} queries. Finally, we will specify ${\mathcal{S}}$ so that it outputs whatever ${\mathcal{A}}$ outputs.

We will prove that  ${\mathcal{S}}$ is such that, given any adversary ${\mathcal{A}}$, any distinguisher ${\mathcal{D}}$ that can distinguish between the outputs $\textsf{out}_{\mathcal{A}}$ in the real world and $\textsf{out}_{\mathcal{S}}$ in the ideal world can be used to build an adversary ${{\mathcal{B}}}$ which breaks the PRF security of $F$. Under the assumption that $F$ is a good PRF, no such efficient distinguisher ${{\mathcal{D}}}$ can exist, and we may conclude that (up to PRF security) ${\mathcal{D}}$ cannot tell that it is interacting with the ideal world instead of the real world.

Why does this proof imply that ${\mathcal{A}}$ cannot manipulate the cardinality estimate? To see this, note that (amongst many other possibilities) ${\mathcal{A}}$'s output could be the vector of bucket values $M[j]$ obtained from a \texttt{corrupt} query, enabling a specific ${\mathcal{D}}$ to compute the cardinality estimate for itself and to attempt to use that estimate as a distinguisher between the two worlds. Yet, by the preceding argument, the simulator in the ideal world is able to produce an output that ${\mathcal{D}}$ cannot distinguish from that of ${\mathcal{A}}$ in the real world. In particular, our simulator will output whatever ${\mathcal{A}}$ outputs, so if ${\mathcal{A}}$ outputs the result of a \texttt{corrupt} query in the ideal world, then so will ${\mathcal{S}}$. But in the ideal world, ${\mathcal{S}}$ uses random sampling for the $y$ values, so there we are operating in the setting where the average case analysis of HLL applies, as per~\cite{hll}. So in the ideal world this specific ${\mathcal{D}}$ would end up computing an accurate average case cardinality estimate. Since ${\mathcal{D}}$ tries to distinguish the two worlds by computing the cardinality estimate and we already showed that no efficient distinguisher ${{\mathcal{D}}}$ can exist, it must follow that the cardinality estimates produced in the real and ideal worlds are indistinguishable (up to PRF security). It then follows that ${\mathcal{A}}$ has no advantage when it comes to manipulating the HLL cardinality estimate. It follows that HLL, when instantiated with PRF $F$, is secure against cardinality manipulation attacks in S4 (and therefore in the weaker attacks models S2 and S3 as well).

@@

We consider distinguisher ${\mathcal{D}}$ to be an algorithm running in a security game. The game begins with the selection of a bit $b$ uniformly at random and then runs either ${\mathcal{A}}$ (the real world, $b=0$) or ${\mathcal{S}}$  (the ideal world, $b=0$). Adversary ${\mathcal{A}}$ interacts with up to four oracles which are handled by the game using the algorithms of HLL-$F$; different combinations of oracles define scenarios S2, S3, S4. After making some queries, ${\mathcal{A}}$ returns its output $\textsf{out}_{\mathcal{A}}$ to ${\mathcal{D}}$. Simulator ${\mathcal{S}}$ runs ${\mathcal{A}}$ in a blackbox manner (as specified below) and receives ${\mathcal{A}}$'s output. It then provides its own output $\textsf{out}_{\mathcal{S}}$ to ${\mathcal{D}}$. Finally, ${\mathcal{D}}$ outputs a bit $b'$ estimating whether it is interacting with the real world ($b'=0$) or the ideal world ($b'=1$). For $i=2, 3, 4$, we define ${\mathcal{D}}$'s S$i$-Real-or-Ideal advantage against HLL-$F$ with adversary ${\mathcal{A}}$ and simulator ${\mathcal{S}}$ to be:
\[
\mbox{Adv}^{\mbox{\tiny S$i$-RoI}}_{\mbox{\tiny{HLL-$F$, ${\mathcal{A}}$, ${\mathcal{S}}$}}}({\mathcal{D}}) := \lvert 2\Pr[b' = b] - 1 \rvert \, .
\]
%Using a standard transformation, this probability can be rewritten as $\lvert \Pr[b' = 1| b = 1] - \Pr[b'=0|b=1] \rvert$.

We say that the scheme HLL-$F$ is S$i$-simulation-secure with parameters $(q,t_{\mathcal{S}},t_{\mathcal{A}},t_{\mathcal{D}},\epsilon)$ if there exists a simulator $\mathcal{S}$ running in time at most $t_{\mathcal{S}}$ such that, for all adversaries ${\mathcal{A}}$ running in time at most $t_{\mathcal{A}}$ and making at most $q$ \texttt{insert($\cdot$)} queries and all distinguishers ${\mathcal{D}}$ running in time at most $t_{\mathcal{D}}$, we have
$\mbox{Adv}^{\mbox{\tiny S$i$-RoI}}_{\mbox{\tiny{HLL-$F$, ${\mathcal{A}}$, ${\mathcal{S}}$}}}({\mathcal{D}}) \leq \epsilon$.

The four oracles that  ${\mathcal{A}}$ can interact with behave as follows in the real world:
\begin{itemize}
\item \texttt{init($\cdot$)}: this oracle receives as input the HLL parameters ($n$, $\ell$, $\alpha_n$) and performs initialisation of the HLL sketch. It also selects a key $K \leftarrow_{\$} \{0,1\}^k$ for keyed function $F: \{0,1\}^k \times {{\mathcal{D}}} \rightarrow  \{0,1\}^{n+\ell}$. This oracle has no output. This oracle is available in all of S2, S3, S4. We assume it is only called once as ${\mathcal{A}}$'s first oracle query.
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$, the oracle implements HLL-$F$ by setting $y = F_K(x)$ and following the procedure in Figure~\ref{fig:hll} to insert value $x$ into the HLL sketch (that is, it uses the first $m$ bits of $y$ to determine a bucket index $j$ and the next $\ell$ bits of $y$ to determine a bit-string $w$; then $\rho(w)$ is used to update the bucket at index $j$ via $M[j]:= \max(M[j],\rho(w))$). This oracle has no output. This oracle is available in all of S2, S3, S4. Without loss of generality (since HLL and variants of it we consider here have a deterministic insertion procedure), we assume all queries $x$ made by ${\mathcal{A}}$ to \texttt{insert($\cdot$)} are distinct.
\item \texttt{read}: this oracle has no input and returns the HLL cardinality estimate as per Figure~\ref{fig:hll}. We allow multiple queries to this oracle and make it available to ${\mathcal{A}}$ in S3 and S4, but not S2.
\item \texttt{corrupt}: this oracle has no input returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents (but not any secret key $K$). This oracle is available only in S4. We allow the adversary to query \texttt{corrupt} as often as it wishes during its attack. Our description of S4 in Section~\ref{sec:attacks} limited the adversary to at most one such query; here we allow multiple queries because doing so does not increase the difficulty of achieving security.  
\end{itemize}

@@

We next describe our simulator ${\mathcal{S}}$. Recall that ${\mathcal{S}}$ has blackbox access to ${\mathcal{A}}$, so we describe how it handles ${\mathcal{A}}$'s queries: 
\begin{itemize}
\item \texttt{init($\cdot$)}: on input $n$, $\ell$, $\alpha_m$, ${\mathcal{S}}$ initialises an HLL sketch as in Figure~\ref{fig:hll}; that is, it creates an array of $m=2^n$ buckets $M[j]$ and initialises them all to 0.
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$,  ${\mathcal{S}}$ ignores $x$, sets $y \leftarrow_{\$} \{0,1\}^{n+\ell}$, and proceeds to update the buckets using this value of $y$ as in the original HLL algorithm in Figure~\ref{fig:hll}.
\item \texttt{read}:  ${\mathcal{S}}$ returns the HLL cardinality estimate computed as per Figure~\ref{fig:hll}.
\item \texttt{corrupt}: ${\mathcal{S}}$ returns the current vector $(M[j])_{j=1}^{m}$ of HLL bucket contents that it has stored.
\end{itemize}
Simulator ${\mathcal{S}}$ operates by running ${\mathcal{A}}$ and handling all of its queries as above. At the end of ${\mathcal{A}}$'s execution, ${\mathcal{S}}$ takes ${\mathcal{A}}$'s output and forwards it to ${\mathcal{D}}$ as its own output.

@@

Now we are ready to formally state and prove our main result:

\todo{pickup from here}

\begin{theorem}\label{thm:security-of-HLL-with-PRF}
Let simulator $\mathcal{S}$ be as described above. Suppose ${\mathcal{A}}$ is any S4 adversary against HLL instantiated with a keyed function $F: \{0,1\}^k \times {{\mathcal{D}}} \rightarrow  \{0,1\}^{n+\ell}$. Suppose ${\mathcal{A}}$ either plays in the real world or in the ideal world with $\mathcal{S}$. Then we can construct from ${\mathcal{A}}$ a PRF adversary ${{\mathcal{B}}}$ against $F$ such that:
\[
 \mbox{Adv}^{\mbox{\tiny PRF}}_{F}({{\mathcal{B}}}) = \mbox{Adv}^{\mbox{\tiny S4-RoI}}_{\mbox{\tiny HLL}}({\mathcal{D}}) \,.
\]
Moreover, if ${\mathcal{A}}$ makes $q$ queries to its \texttt{insert($\cdot$)} oracle, then ${{\mathcal{B}}}$ makes $q$ queries to its real or random oracle. Finally, the running time of ${{\mathcal{B}}}$ is (roughly) the same as that of ${\mathcal{A}}$.
\end{theorem}

@@

\begin{proof}
PRF adversary ${{\mathcal{B}}}$ against $F$ is built as follows. Let $d$ denote the hidden bit determining whether ${{\mathcal{B}}}$ receives real responses of the form $F_K(x)$ where $K$ is a uniformly random key ($d=0$) or responses from a truly random function ($d=1$); ${{\mathcal{B}}}$ runs S4 adversary ${\mathcal{A}}$ and responds to each of its queries as follows:
\begin{itemize}
\item \texttt{init}: here ${{\mathcal{B}}}$ initialises an HLL sketch in accordance with Figure~\ref{fig:hll}. Thus it initialises an array $M[j]$ in which each entry is set to 0. 
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$,  ${{\mathcal{B}}}$ makes a query on $x$ to its real-or-random oracle and receives value $y$; ${{\mathcal{B}}}$ uses $y$ to update the HLL bucket $j$ generated via $j=1 + <y_1y_2...y_{n}>_2$ as in the regular HLL algorithm (shown in Figure~\ref{fig:hll}).
\item \texttt{read}: ${{\mathcal{B}}}$ uses its knowledge of the buckets' contents to compute and return the HLL cardinality estimate in accordance with Figure~\ref{fig:hll}. 
\item \texttt{corrupt}: ${{\mathcal{B}}}$ returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents. 
\end{itemize}
When ${\mathcal{A}}$ outputs a bit $b'$, ${{\mathcal{B}}}$ outputs its own bit $d'=b'$. Note that when $d=0$, ${{\mathcal{B}}}$ plays against the real $F$ with a random key $K$ and provides a simulation for ${\mathcal{A}}$ that perfectly instantiates the real world. On the other hand, when $d=1$, ${{\mathcal{B}}}$ plays against a truly random function and provides a simulation for ${\mathcal{A}}$ that perfectly instantiates the ideal world with simulator $\mathcal{S}$ (to see this, note that sampling the outputs $y$ of a truly random function at distinct inputs $x$ is identical to simply sampling $y$ uniformly at random). The equality of advantages in the statement of the theorem now follows immediately. The remainder of the proof follows on counting queries and noting that the HLL algorithm for computing cardinality is efficient, so ${{\mathcal{B}}}$'s running time due to handling \texttt{read} queries is not increased significantly compared to that of ${\mathcal{A}}$.
\end{proof}

We reiterate our central point: the theorem shows that no efficient S4 adversary ${\mathcal{A}}$ can tell that it is playing against our simulator ${\mathcal{S}}$ using truly random values $y$ instead of with values obtained from a PRF, except with a small advantage. This means that ${\mathcal{A}}$ is effectively reduced to attacking HLL in the exact same setting used in the average case analysis of HLL from~\cite{hll}, wherein it is proved that the HLL cardinality estimate endures relative errors of the form $1.04/\sqrt{m}$. From this we can conclude that no efficient S4 adversary ${\mathcal{A}}$ can manipulate the HLL cardinality estimate to its benefit.

Note that our security analysis using a simulation-based definition does not require us to specify a winning condition for the adversary. Rather, we infer from inspecting the simulator what the adversary can (or cannot) achieve in the ideal world, and show that no efficient adversary can distinguish the real and ideal worlds. This stands in contrast to the game-based approach of~\cite{CCS:ClaPatShrPS19} which needs a complex definition to say when the adversary has successfully manipulated a probabilistic data structure. This reveals a strength of the simulation-based approach in more complex settings like ours.

@@

An HLL can leak information about values $x$ that were previously inserted. Consider the following attack from~\cite{cardestprivacy}: to test whether a value $x$ was previously inserted into the HLL by an honest user (say), the adversary makes the sequence of queries: \texttt{read}, \texttt{insert($x$)}, \texttt{read}. If the two cardinality estimates are different, the adversary can conclude definitively that $x$ was \emph{not} previously inserted into the HLL; if they are the same, it can estimate (with some error) that $x$ was previously inserted.\footnote{Errors can arise in two ways when $x$ was not previously inserted: either inserting $x$ changes a bucket value, but the change is masked by rounding when the cardinality estimate is computed, or inserting $x$ does lead to a change in a bucket value because $\rho(w)$ computed from $x$ is too small.} This attack still works even if HLL is implemented using a PRF and is unavoidable in view of the oracle queries (and in real settings, the API) available to the adversary. So such privacy leaks are an inherent feature of HLL if a suitably rich API is provided. Note that this attack does not contradict our main result, which concerns prevention of cardinality manipulation rather than confidentiality.

Still, simulation-based security notions are often used to model confidentiality properties, so it is tempting to think that our security definition might say something about confidentiality of the HLL inputs $x$. Our existing model does not, since ${\mathcal{A}}$ chooses all the inputs $x$ and so ``knows'' them already. However, it is easily modified to provide a confidentiality notion. For example, we can consider the insertion of honestly generated values $x$ that are drawn from some probability distribution $p$ on $\mathcal{D}$ in a first phase, with ${\mathcal{A}}$ then having access to only \texttt{read} and \texttt{corrupt} oracles in a second phase (we have to remove the \texttt{insert($\cdot$)} oracle in view of the general attack from~\cite{cardestprivacy}).

In this case, the simulator ${\mathcal{S}}$ in the ideal world can proceed without ever seeing any inputs but instead knowing only the distribution $p$ and the number of distinct samples drawn from it in the first phase. It is not hard to modify the proof of Theorem~\ref{thm:security-of-HLL-with-PRF} to show that no adversary can distinguish real and ideal worlds in this case (up to PRF security). Now, since ${\mathcal{S}}$ has no access to inputs $x$, we can conclude that the HLL implemented using a PRF ensures confidentiality for the inputs $x$. Thus, the cardinality estimates and even the bucket values themselves do not leak information about the honestly inserted values $x$ (beyond their number).

We leave the further exploration of how to adapt our simulation-based security definitions to capture confidentiality notions to future work.

>>>>

\todo[inline]{This ends up being pretty harsh on this NSDI paper, but hey. We may shorten or move some details to an appendix.}

@@

\subsection{Intuition for Formal Security Analysis}

@@

In the ideal world ($b=1$), ${\mathcal{A}}$ is replaced by a simulator ${\mathcal{S}}$ with blackbox oracle access to ${\mathcal{A}}$ (that is, ${\mathcal{S}}$ can run ${\mathcal{A}}$ in a blackbox manner, receiving its oracle queries and providing responses to ${\mathcal{A}}$ as it sees fit). In particular, ${\mathcal{S}}$ receives parameters $n$, $\ell$, $\alpha_n$ from ${\mathcal{A}}$'s \texttt{init($\cdot$)} call. Finally, ${\mathcal{S}}$ receives $\textsf{out}_{\mathcal{A}}$ and completes its execution by producing its own output $\textsf{out}_{\mathcal{S}}$.

Importantly, we will specify an ${\mathcal{S}}$ that handles all of ${\mathcal{A}}$'s \texttt{insert($x$)} queries using \emph{random sampling}, that is, by setting $y \leftarrow_{\$} \{0,1\}^{n+\ell}$ instead of $y: = F_K(x)$ and then executing the rest of the insertion process as in Figure~\ref{fig:hll}. In essence, our ${\mathcal{S}}$ will build its own ``shadow'' HLL in which all insertion queries are handled by random sampling. ${\mathcal{S}}$ then uses the shadow HLL for handling ${\mathcal{A}}$'s \texttt{read} and \texttt{corrupt} queries. Finally, we will specify ${\mathcal{S}}$ so that it outputs whatever ${\mathcal{A}}$ outputs.

We will prove that  ${\mathcal{S}}$ is such that, given any adversary ${\mathcal{A}}$, then any distinguisher $\mathcal{E}$ that can distinguish between the $\textsf{out}_{\mathcal{A}}$ in the real world and $\textsf{out}_{\mathcal{S}}$ in the ideal world can be used to build an adversary ${{\mathcal{B}}}$ which breaks the PRF security of $F$. Under the assumption that $F$ is a good PRF, no such efficient distinguisher $\mathcal{E}$ can exist, and we may conclude that (up to PRF security) ${\mathcal{E}}$ cannot tell that it is interacting with the ideal world instead of the real world.

Why does this proof imply that ${\mathcal{A}}$ cannot manipulate the cardinality estimate? To see this, note that (amongst many other possibilities) ${\mathcal{A}}$'s output could be the vector of bucket values $M[j]$ obtained from a \texttt{corrupt} query, enabling a specific ${\mathcal{E}}$ to compute the cardinality estimate for itself and to attempt to use that estimate as a distinguisher between the two worlds. Yet, by the preceding argument, the simulator in the ideal world is able to produce an output that ${\mathcal{E}}$ cannot distinguish from that of ${\mathcal{A}}$ in the real world. In particular, our simulator will output whatever ${\mathcal{A}}$ outputs, so if ${\mathcal{A}}$ outputs the result of a \texttt{corrupt} query in the ideal world, then so will ${\mathcal{S}}$. But in the ideal world, ${\mathcal{S}}$ uses random sampling for the $y$ values, so there ${\mathcal{S}}$ (and ${\mathcal{A}}$ that it runs) are operating in the setting where the average case analysis of HLL applies, as per~\cite{hll}. So in the ideal world this specific ${\mathcal{E}}$ would end up computing an accurate average case cardinality estimate. Since ${\mathcal{E}}$ tries to distinguish the two worlds by computing the cardinality estimate and we already showed that no efficient distinguisher ${{\mathcal{E}}}$ can exist, it must follow that the cardinality estimates produced in the real and ideal worlds are indistinguishable (up to PRF security). It then follows that ${\mathcal{A}}$ has no advantage when it comes to manipulating the HLL-$F$ cardinality estimate. So HLL, when instantiated with PRF $F$, is secure against cardinality manipulation attacks in S4 (and therefore in the weaker attacks models S2 and S3 as well).

@@

We now formalise the above intuition. We consider distinguisher ${\mathcal{E}}$ to be an algorithm running in a security game. The game begins with the selection of a bit $b$ uniformly at random and then runs either ${\mathcal{A}}$ (the real world, $b=0$) or ${\mathcal{S}}$ (the ideal world, $b=1$). Adversary ${\mathcal{A}}$ interacts with up to four oracles which are handled by the game using the algorithms of HLL-$F$; different combinations of oracles define scenarios S2, S3, S4. After making some queries, ${\mathcal{A}}$ returns its output $\textsf{out}_{\mathcal{A}}$ to ${\mathcal{E}}$. Simulator ${\mathcal{S}}$ runs ${\mathcal{A}}$ in a blackbox manner (as specified below) and receives ${\mathcal{A}}$'s output. It then provides its own output $\textsf{out}_{\mathcal{S}}$ to ${\mathcal{E}}$. Finally, ${\mathcal{E}}$ outputs a bit $b'$ estimating whether it is interacting with the real world ($b'=0$) or the ideal world ($b'=1$). For $i=2, 3, 4$, we define ${\mathcal{E}}$'s S$i$-Real-or-Ideal advantage against HLL-$F$ with adversary ${\mathcal{A}}$ and simulator ${\mathcal{S}}$ to be:
\[
\mbox{Adv}^{\mbox{\tiny S$i$-RoI}}_{\mbox{\tiny{HLL-$F$, ${\mathcal{A}}$, ${\mathcal{S}}$}}}({\mathcal{E}}) := \lvert 2\Pr[b' = b] - 1 \rvert \, .
\]
%Using a standard transformation, this probability can be rewritten as $\lvert \Pr[b' = 1| b = 1] - \Pr[b'=0|b=1] \rvert$.

We say that the scheme HLL-$F$ is S$i$-simulation-secure with parameters $(q,t_{\mathcal{S}},t_{\mathcal{A}},t_{\mathcal{E}},\epsilon)$ if there exists a simulator $\mathcal{S}$ running in time at most $t_{\mathcal{S}}$ such that, for all adversaries ${\mathcal{A}}$ running in time at most $t_{\mathcal{A}}$ and making at most $q$ \texttt{insert($\cdot$)} queries and all distinguishers ${\mathcal{E}}$ running in time at most $t_{\mathcal{E}}$, we have
$\mbox{Adv}^{\mbox{\tiny S$i$-RoI}}_{\mbox{\tiny{HLL-$F$, ${\mathcal{A}}$, ${\mathcal{S}}$}}}({\mathcal{E}}) \leq \epsilon$. We quantify over \texttt{insert($\cdot$)} queries (via $q$) and not other types of query because only \texttt{insert($\cdot$)} queries will appear in our concrete security analysis and we do not want to clutter the notation.

The four oracles that  ${\mathcal{A}}$ can interact with behave as follows in the real world:
\begin{itemize}
\item \texttt{init($\cdot$)}: this oracle receives as input the HLL-$F$ parameters ($n$, $\ell$, $\alpha_n$) and performs initialisation of the HLL-$F$ sketch, including generating the array of buckets $M[j]$ and initialising all the entries to $0$, and selecting a key $K \leftarrow_{\$} \{0,1\}^k$ for keyed function $F: \{0,1\}^k \times {{\mathcal{D}}} \rightarrow  \{0,1\}^{n+\ell}$. This oracle has no output. It is available in all of S2, S3, S4. We assume it is only called once as ${\mathcal{A}}$'s first oracle query.
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$, the oracle implements HLL-$F$ by setting $y = F_K(x)$ and following the procedure in Figure~\ref{fig:hll} to insert value $x$ into the HLL sketch (that is, it uses the first $m$ bits of $y$ to determine a bucket index $j$ and the next $\ell$ bits of $y$ to determine a bit-string $w$; then $\rho(w)$ is used to update the bucket at index $j$ via $M[j]:= \max(M[j],\rho(w))$). This oracle has no output. It is available in all of S2, S3, S4. Without loss of generality (since HLL and variants of it we consider here have a deterministic insertion procedure), we assume all queries $x$ made by ${\mathcal{A}}$ to \texttt{insert($\cdot$)} are distinct.
\item \texttt{read}: this oracle has no input and returns the HLL-$F$ cardinality estimate as per Figure~\ref{fig:hll}, i.e.\ it computes a raw cardinality estimate via $E := \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$ and then performs corrections for small and large values. We allow multiple queries to this oracle and make it available to ${\mathcal{A}}$ in S3 and S4, but not S2.
\item \texttt{corrupt}: this oracle has no input and returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents (but not any secret key $K$). This oracle is available only in S4. We allow the adversary to query \texttt{corrupt} as often as it wishes during its attack. Our description of S4 in Section~\ref{sec:attacks} limited the adversary to at most one such query; here we allow multiple queries because doing so does not increase the difficulty of achieving security.  
\end{itemize}

@@

We next describe the simulator ${\mathcal{S}}$ for the ideal world. Simulator ${\mathcal{S}}$ operates by running ${\mathcal{A}}$ in a blackbox manner and handling all of its queries as follows: 
\begin{itemize}
\item \texttt{init($\cdot$)}: on input $n$, $\ell$, $\alpha_m$, ${\mathcal{S}}$ initialises an HLL sketch as in Figure~\ref{fig:hll}; that is, it creates an array of $m=2^n$ buckets $M[j]$ and initialises them all to 0.
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$,  ${\mathcal{S}}$ ignores $x$, sets $y \leftarrow_{\$} \{0,1\}^{n+\ell}$, and proceeds to update the buckets using this value of $y$ as in the original HLL algorithm in Figure~\ref{fig:hll}.
\item \texttt{read}:  ${\mathcal{S}}$ returns the HLL cardinality estimate computed as per Figure~\ref{fig:hll}.
\item \texttt{corrupt}: ${\mathcal{S}}$ returns the current vector $(M[j])_{j=1}^{m}$ of HLL bucket contents that it has stored.
\end{itemize}
At the end of ${\mathcal{A}}$'s execution, ${\mathcal{S}}$ takes ${\mathcal{A}}$'s output and forwards it to ${\mathcal{E}}$ as its own output. Notice that the running time of  ${\mathcal{S}}$ is (roughly) equal to that of ${\mathcal{A}}$.

@@

Now we are ready to formally state and prove our main result:
\begin{theorem}\label{thm:security-of-HLL-with-PRF}
Let simulator $\mathcal{S}$ be as described above. For any S4 adversary ${\mathcal{A}}$ against HLL-$F$ and any distinguisher $\mathcal{E}$, we can construct a PRF adversary ${{\mathcal{B}}}$ against $F$ such that:
\[
 \mbox{Adv}^{\mbox{\tiny PRF}}_{F}({{\mathcal{B}}}) = 
 \mbox{Adv}^{\mbox{\tiny S$4$-RoI}}_{\mbox{\tiny{HLL-$F$, ${\mathcal{A}}$, ${\mathcal{S}}$}}}({\mathcal{E}}) \,.
\]
Moreover, if ${\mathcal{A}}$ makes $q$ queries to its \texttt{insert($\cdot$)} oracle, then ${{\mathcal{B}}}$ makes $q$ queries to its real or random oracle. Finally, the running time of ${{\mathcal{B}}}$ is (roughly) $t_{\mathcal{A}}+t_{\mathcal{E}}$.
\end{theorem}

@@

\begin{proof}
PRF adversary ${{\mathcal{B}}}$ against $F$ is built as follows. Let $d$ denote the hidden bit determining whether ${{\mathcal{B}}}$ receives real responses of the form $F_K(x)$ where $K$ is a uniformly random key ($d=0$) or responses from a truly random function ($d=1$); ${{\mathcal{B}}}$ provides a simulation of either the real or random world for distinguisher ${\mathcal{E}}$ as follows. ${{\mathcal{B}}}$ runs ${{\mathcal{A}}}$ and answers each of ${{\mathcal{A}}}$'s queries as follows:
\begin{itemize}
\item \texttt{init($\cdot$)}: ${{\mathcal{B}}}$ receives parameters $n$, $\ell$, $\alpha_m$ and initialises an array $M[j]$ in which each of the $2^n$ entries is set to 0. 
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$,  ${{\mathcal{B}}}$ makes a query on $x$ to its real-or-random oracle and receives value $y$; ${{\mathcal{B}}}$ uses $y$ to update the HLL sketch as in the regular HLL algorithm (shown in Figure~\ref{fig:hll}). That is, it uses the first $m$ bits of $y$ to determine a bucket index $j$ and the next $\ell$ bits of $y$ to determine a bit-string $w$; then $\rho(w)$ is used to update the bucket at index $j$ via $M[j]:= \max(M[j],\rho(w))$.
\item \texttt{read}: ${{\mathcal{B}}}$ uses its knowledge of the buckets' contents to compute and return the HLL cardinality estimate as per Figure~\ref{fig:hll}. 
\item \texttt{corrupt}: ${{\mathcal{B}}}$ returns the vector $(M[j])_{j=1}^{m}$. 
\end{itemize}
When ${\mathcal{A}}$ returns its final output $\textsf{out}_{\mathcal{S}}$, ${{\mathcal{B}}}$ forwards this output to ${{\mathcal{E}}}$. Finally, ${{\mathcal{E}}}$ outputs a bit $b'$. Then ${{\mathcal{B}}}$ outputs its own bit $d'=b'$.

Note that when $d=0$, ${{\mathcal{B}}}$ plays against the real $F$ with a random key $K$ and provides a simulation for ${\mathcal{A}}$ that perfectly instantiates the real world. On the other hand, when $d=1$, ${{\mathcal{B}}}$ plays against a truly random function and perfectly instantiates our simulator $\mathcal{S}$ in the ideal world. To see this, note that sampling the outputs $y$ of a truly random function at distinct inputs $x$ is identical to simply sampling $y$ uniformly at random.\footnote{Our proof relies on the $x$ being distinct, but this is just to simplify book-keeping and is without loss of generality.} The equality of advantages in the statement of the theorem now follows immediately. The remainder of the proof follows on counting queries and noting that the HLL sub-algorithms for insertion and computing cardinality are efficient, so ${{\mathcal{B}}}$'s running time due to handling ${\mathcal{A}}$'s queries is not increased significantly compared to that of ${\mathcal{A}}$. The final running time for ${\mathcal{B}}$, namely $t_{\mathcal{A}}+t_{\mathcal{E}}$, follows on noting that ${\mathcal{B}}$ first runs ${\mathcal{A}}$ and then ${\mathcal{E}}$.
\end{proof}

Note that our security analysis using a simulation-based definition does not require us to specify a winning condition for the adversary. Rather, we infer from inspecting the simulator what the adversary can (or cannot) achieve in the ideal world, and show that no efficient adversary can distinguish the real and ideal worlds. This is in contrast to the game-based approach of~\cite{CCS:ClaPatShrPS19} which needs a complex definition to say when the adversary has successfully manipulated a probabilistic data structure. This reveals a strength of the simulation-based approach in more complex settings like ours.

@@

An HLL can leak information about values $x$ that were previously inserted. Consider the following attack from~\cite{cardestprivacy}: to test whether a value $x$ was previously inserted into the HLL by an honest user (say), the adversary makes the sequence of queries: \texttt{read}, \texttt{insert($x$)}, \texttt{read}. If the two cardinality estimates are different, the adversary can conclude definitively that $x$ was \emph{not} previously inserted into the HLL; if they are the same, it can estimate (with some error) that $x$ was previously inserted.\footnote{Errors can arise in two ways when $x$ was not previously inserted: either inserting $x$ changes a bucket value, but the change is masked by rounding when the cardinality estimate is computed, or inserting $x$ does lead to a change in a bucket value because $\rho(w)$ computed from $x$ is too small.} This attack still works even if HLL is implemented using a PRF and is unavoidable in view of the oracle queries (and in real settings, the API) available to the adversary. So such privacy leaks are an inherent feature of HLL if a rich enough API is provided. Note that this attack does not contradict our main result, which concerns prevention of cardinality manipulation rather than confidentiality.

%Still, simulation-based security notions are often used to model confidentiality properties, so it is tempting to think that our security definition might say something about confidentiality of the HLL inputs $x$. Our existing model does not, since ${\mathcal{A}}$ chooses all the inputs $x$ and so ``knows'' them already. However, it is easily modified to provide a confidentiality notion. For example, we can consider the insertion of honestly generated values $x$ that are drawn from some probability distribution $p$ on $\mathcal{D}$ in a first phase, with ${\mathcal{A}}$ then having access to only \texttt{read} and \texttt{corrupt} oracles in a second phase (we have to remove the \texttt{insert($\cdot$)} oracle in view of the general attack from~\cite{cardestprivacy}).

%In this case, the simulator ${\mathcal{S}}$ in the ideal world can proceed without ever seeing any inputs but instead knowing only the distribution $p$ and the number of distinct samples drawn from it in the first phase. It is not hard to modify the proof of Theorem~\ref{thm:security-of-HLL-with-PRF} to show that no adversary can distinguish real and ideal worlds in this case (up to PRF security). Now, since ${\mathcal{S}}$ has no access to inputs $x$, we can conclude that the HLL implemented using a PRF ensures confidentiality for the inputs $x$. Thus, the cardinality estimates and even the bucket values themselves do not leak information about the honestly inserted values $x$ (beyond their number).

We leave the exploration of the adaptation of our simulation-based security definitions to capture confidentiality notions for more restrictred APIs to future work.
